---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mtuem

<!-- badges: start -->
<!-- badges: end -->

The **mtuem** package implements the likelihood functions for estimating Microeconomic Time-Use-Expenditure Models (MTUEM) within the [Apollo](https://www.apollochoicemodelling.com/) choice modelling framework. The models are based on Jara-Diaz et al. (2008), which maximizes a Cobb-Douglas utility function over time-use and goods consumption subject to time, monetary, and technical constraints.

## Installation

```r
install.packages("devtools")
devtools::install_github("https://github.com/pabloreyespo/mtuem")
```

## Model Formulation Guide

### The MTUEM Framework

The MTUEM derives optimal allocations of time and expenditure from utility maximization. The model has two budget constraints:

- **Time budget**: Total time equals the time budget (typically 168 hours per week), split into committed time (Tc), work time (Tw), and freely allocated time.
- **Monetary budget**: Total expenditure equals labor income minus committed expenses (Ec), split into freely allocated goods.

### Two Parameterizations

The package offers two equivalent parameterizations:

| | Theta-Phi (`mtuem_likelihood`) | Alpha-Beta (`mtuem_ab_likelihood`) |
|---|---|---|
| Work elasticities | `Theta`, `Phi`, `thw` | `alpha`, `beta` |
| Identification | Fix `Theta = 1` | No fixed parameters needed |
| Use when | Direct interpretation of time value | Alternative formulation |

### Equations and the "Leave One Out" Rule

- **work_times** (required): Column name(s) for work time, typically `c("Tw")`.
- **free_times** (optional): Column names for freely allocated time activities. You must leave at least one out for identification. The omitted category is recovered as the residual from the time budget.
- **free_goods** (optional): Column names for freely allocated goods. You must leave at least one out for identification. The omitted category is recovered as the residual from the monetary budget.

### Error Structure

The likelihood assumes normally distributed errors. Three specification options exist:

1. **Inferred covariance** (no `sig`, no `rho`): The error covariance is inferred from residuals. Simplest approach, valid only for single-equation models.
2. **Explicit sigma, inferred correlations** (`sig` specified, no `rho`): Standard deviations are estimated as parameters. Correlations are inferred from residuals. Valid for multi-equation models.
3. **Full estimation** (both `sig` and `rho` specified): Both standard deviations and correlations are estimated as parameters.

### Retrieving the Covariance Matrix

After estimation, use `mtuem_get_corr(model, apollo_probabilities, apollo_inputs, vars)` where `vars` is a character vector of the time-use-expenditure column names.

---

## Example 1: Single Equation (Work Only)

The simplest model estimates only the work equation. No free times or free goods are specified.

```{r example-1eq, eval = FALSE}
library(mtuem)
library(apollo)

### Initialise
apollo_initialise()
apollo_control <- list(
  modelName = "maed-1eq",
  modelDescr = "Single equation MTUEM (work only)",
  indivID = "PeID",
  outputDirectory = "output"
)

### Load data
database <- mtuem::maed
database <- database[database$EcI > 0, ]

### Parameters: Theta fixed to 1 for identification
apollo_beta <- c(Theta = 1, Phi = 1, thw = 0)
apollo_fixed <- c("Theta")

apollo_inputs <- apollo_validateInputs()

apollo_probabilities <- function(apollo_beta, apollo_inputs, functionality = "estimate") {
  apollo_attach(apollo_beta, apollo_inputs)
  on.exit(apollo_detach(apollo_beta, apollo_inputs))

  P <- list()

  work_elasticities <- list(Theta = Theta, Phi = Phi, thw = thw)

  mtuem_settings <- list(
    work_times = c("Tw"),
    work_elasticities = work_elasticities,
    Tc = Tc, Ec = EcI, w = w, tau = 168
  )

  P[["model"]] <- mtuem_likelihood(mtuem_settings, functionality)
  P <- apollo_prepareProb(P, apollo_inputs, functionality)
  return(P)
}

model <- apollo_estimate(apollo_beta, apollo_fixed,
                          apollo_probabilities, apollo_inputs)
apollo_modelOutput(model)

### Predictions
pred <- apollo_prediction(model, apollo_probabilities, apollo_inputs)
head(pred)

### Values of Time
Tw <- pred$Tw
w  <- database$w
Tc <- database$Tc
Ec <- database$Ec

cteVoL  <- mean((w * Tw - Ec) / (168 - Tw - Tc))
cteVTAW <- mean((w * Tw - Ec) / Tw)

delta <- apollo_deltaMethod(model, list(
  expression = c(
    VoL  = paste0(cteVoL,  "*Theta/Phi"),
    VTAW = paste0(cteVTAW, "*thw/Phi")
  )
))
delta
```

---

## Example 2: Three Equations with Different Error Structures

This example shows a model with 3 equations: work (Tw), one free time (Tf1), and one free good (Ef1). We demonstrate three error structure specifications.

### 2a: Inferred Covariance Matrix

No `sig` or `rho` parameters. The covariance is inferred from residuals.

```{r example-3eq-inferred, eval = FALSE}
library(mtuem)
library(apollo)

apollo_initialise()
apollo_control <- list(
  modelName = "maed-3eq-inferred",
  modelDescr = "Three equations with inferred covariance",
  indivID = "PeID",
  outputDirectory = "output"
)

database <- mtuem::maed
database <- database[database$EcI > 0, ]

### Only structural parameters; no sig or rho
apollo_beta <- c(Theta = 1, Phi = 1, thw = 0,
                 theta_1 = 0.5, phi_1 = 0.5)
apollo_fixed <- c("Theta")

apollo_inputs <- apollo_validateInputs()

apollo_probabilities <- function(apollo_beta, apollo_inputs, functionality = "estimate") {
  apollo_attach(apollo_beta, apollo_inputs)
  on.exit(apollo_detach(apollo_beta, apollo_inputs))

  P <- list()

  work_elasticities <- list(Theta = Theta, Phi = Phi, thw = thw)
  times_elasticities <- list(theta_1 = theta_1)
  goods_elasticities <- list(phi_1 = phi_1)

  mtuem_settings <- list(
    work_times = c("Tw"),
    free_times = c("Tf1"),
    free_goods = c("Ef1"),
    goods_cost = list(Ef1 = 1),
    work_elasticities = work_elasticities,
    times_elasticities = times_elasticities,
    goods_elasticities = goods_elasticities,
    Tc = Tc, Ec = EcI, w = w, tau = 168
  )

  P[["model"]] <- mtuem_likelihood(mtuem_settings, functionality)
  P <- apollo_prepareProb(P, apollo_inputs, functionality)
  return(P)
}

model <- apollo_estimate(apollo_beta, apollo_fixed,
                          apollo_probabilities, apollo_inputs)

### Retrieve covariance matrix after estimation
corr_result <- mtuem_get_corr(model, apollo_probabilities, apollo_inputs,
                               vars = c("Tw", "Tf1", "Ef1"))
print(corr_result$covar)
print(corr_result$corr)
```

### 2b: Explicit Standard Deviations, Inferred Correlations

Estimate `sig` parameters for each equation. Correlations are inferred from residuals.

```{r example-3eq-sig, eval = FALSE}
library(mtuem)
library(apollo)

apollo_initialise()
apollo_control <- list(
  modelName = "maed-3eq-sig",
  modelDescr = "Three equations with explicit sigma, inferred correlations",
  indivID = "PeID",
  outputDirectory = "output"
)

database <- mtuem::maed
database <- database[database$EcI > 0, ]

### Structural parameters + one sigma per equation
apollo_beta <- c(Theta = 1, Phi = 1, thw = 0,
                 theta_1 = 0.5, phi_1 = 0.5,
                 sig_1 = 100, sig_2 = 100, sig_3 = 100)
apollo_fixed <- c("Theta")

apollo_inputs <- apollo_validateInputs()

apollo_probabilities <- function(apollo_beta, apollo_inputs, functionality = "estimate") {
  apollo_attach(apollo_beta, apollo_inputs)
  on.exit(apollo_detach(apollo_beta, apollo_inputs))

  P <- list()

  work_elasticities <- list(Theta = Theta, Phi = Phi, thw = thw)
  times_elasticities <- list(theta_1 = theta_1)
  goods_elasticities <- list(phi_1 = phi_1)
  sig <- list(sig_1 = sig_1, sig_2 = sig_2, sig_3 = sig_3)

  mtuem_settings <- list(
    work_times = c("Tw"),
    free_times = c("Tf1"),
    free_goods = c("Ef1"),
    goods_cost = list(Ef1 = 1),
    work_elasticities = work_elasticities,
    times_elasticities = times_elasticities,
    goods_elasticities = goods_elasticities,
    sig = sig,
    Tc = Tc, Ec = EcI, w = w, tau = 168
  )

  P[["model"]] <- mtuem_likelihood(mtuem_settings, functionality)
  P <- apollo_prepareProb(P, apollo_inputs, functionality)
  return(P)
}

model <- apollo_estimate(apollo_beta, apollo_fixed,
                          apollo_probabilities, apollo_inputs)
```

### 2c: Full Estimation (Sigma and Rho)

Estimate both standard deviations and correlations. For 3 equations, the upper triangle of the correlation matrix has 3 elements: `rho_12`, `rho_13`, `rho_23`.

```{r example-3eq-full, eval = FALSE}
library(mtuem)
library(apollo)

apollo_initialise()
apollo_control <- list(
  modelName = "maed-3eq-full",
  modelDescr = "Three equations with full covariance estimation",
  indivID = "PeID",
  outputDirectory = "output"
)

database <- mtuem::maed
database <- database[database$EcI > 0, ]

### Structural + sigma + rho parameters
apollo_beta <- c(
  Theta = 1, Phi = 1, thw = 0,
  theta_1 = 0.5, phi_1 = 0.5,
  sig_1 = 100, sig_2 = 100, sig_3 = 100,
  rho_12 = 0, rho_13 = 0, rho_23 = 0
)
apollo_fixed <- c("Theta")

apollo_inputs <- apollo_validateInputs()

apollo_probabilities <- function(apollo_beta, apollo_inputs, functionality = "estimate") {
  apollo_attach(apollo_beta, apollo_inputs)
  on.exit(apollo_detach(apollo_beta, apollo_inputs))

  P <- list()

  work_elasticities <- list(Theta = Theta, Phi = Phi, thw = thw)
  times_elasticities <- list(theta_1 = theta_1)
  goods_elasticities <- list(phi_1 = phi_1)
  sig <- list(sig_1 = sig_1, sig_2 = sig_2, sig_3 = sig_3)
  rho <- list(rho_12 = rho_12, rho_13 = rho_13, rho_23 = rho_23)

  mtuem_settings <- list(
    work_times = c("Tw"),
    free_times = c("Tf1"),
    free_goods = c("Ef1"),
    goods_cost = list(Ef1 = 1),
    work_elasticities = work_elasticities,
    times_elasticities = times_elasticities,
    goods_elasticities = goods_elasticities,
    sig = sig,
    rho = rho,
    Tc = Tc, Ec = EcI, w = w, tau = 168
  )

  P[["model"]] <- mtuem_likelihood(mtuem_settings, functionality)
  P <- apollo_prepareProb(P, apollo_inputs, functionality)
  return(P)
}

model <- apollo_estimate(apollo_beta, apollo_fixed,
                          apollo_probabilities, apollo_inputs)

### Retrieve the estimated covariance matrix
corr_result <- mtuem_get_corr(model, apollo_probabilities, apollo_inputs,
                               vars = c("Tw", "Tf1", "Ef1"))
print(corr_result$covar)
print(corr_result$corr)
```

---

## Example 3: Latent Class Model (2 Classes, 3 Equations each)

This example shows a latent class model with 2 classes, each with its own work equation, 2 free time equations, and full covariance estimation. Class membership is modeled with covariates.

```{r example-lc-2class, eval = FALSE}
library(mtuem)
library(apollo)

est_set <- list(
  writeIter = FALSE,
  silent = TRUE,
  maxIterations = 500,
  scaleHessian = FALSE,
  scaleAfterConvergence = FALSE,
  estimationRoutine = "bfgs",
  hessianRoutine = "maxLik"
)

apollo_initialise()
apollo_control <- list(
  modelName = "maed-lc-2class-3eq",
  modelDescr = "Latent class model with 2 classes and 3 equations each",
  indivID = "PeID",
  outputDirectory = "output",
  noValidation = TRUE,
  noDiagnostics = TRUE
)

database <- mtuem::maed
database <- database[database$EcI > 0, ]

### Parameters for 2 classes
apollo_beta <- c(
  # Class 1 structural
  Theta_1 = 1, Phi_1 = 1, thw_1 = 0,
  th1_1 = 0.3, th2_1 = 0.3,
  # Class 1 error structure (3 equations: Tw, Tf1, Tf2)
  sig1_1 = 10, sig2_1 = 5, sig3_1 = 5,
  rho12_1 = 0, rho13_1 = 0, rho23_1 = 0,
  # Class 1 membership
  asc_1 = 0, x_female_1 = 0, x_older45_1 = 0,
  # Class 2 structural
  Theta_2 = 1, Phi_2 = 1, thw_2 = 0,
  th1_2 = 0.3, th2_2 = 0.3,
  # Class 2 error structure
  sig1_2 = 10, sig2_2 = 5, sig3_2 = 5,
  rho12_2 = 0, rho13_2 = 0, rho23_2 = 0,
  # Class 2 membership
  asc_2 = 0, x_female_2 = 0, x_older45_2 = 0
)

apollo_fixed <- c(
  "Theta_1", "Theta_2",
  "asc_1", "x_female_1", "x_older45_1"
)

### Class membership parameters
apollo_lcPars <- function(apollo_beta, apollo_inputs) {
  lcpars <- list()
  lcpars[["Theta"]] <- list(Theta_1, Theta_2)
  lcpars[["Phi"]] <- list(Phi_1, Phi_2)
  lcpars[["thw"]] <- list(thw_1, thw_2)
  lcpars[["th1"]] <- list(th1_1, th1_2)
  lcpars[["th2"]] <- list(th2_1, th2_2)
  lcpars[["sig1"]] <- list(sig1_1, sig1_2)
  lcpars[["sig2"]] <- list(sig2_1, sig2_2)
  lcpars[["sig3"]] <- list(sig3_1, sig3_2)
  lcpars[["rho12"]] <- list(rho12_1, rho12_2)
  lcpars[["rho13"]] <- list(rho13_1, rho13_2)
  lcpars[["rho23"]] <- list(rho23_1, rho23_2)

  V <- list()
  V[["class_1"]] <- asc_1 + (female * x_female_1) + (older45 * x_older45_1)
  V[["class_2"]] <- asc_2 + (female * x_female_2) + (older45 * x_older45_2)

  classAlloc_settings <- list(
    classes = c(class_1 = 1, class_2 = 2),
    utilities = V
  )
  lcpars[["pi_values"]] <- apollo_classAlloc(classAlloc_settings)
  return(lcpars)
}

apollo_inputs <- apollo_validateInputs()

apollo_probabilities <- function(apollo_beta, apollo_inputs, functionality = "estimate") {
  apollo_attach(apollo_beta, apollo_inputs)
  on.exit(apollo_detach(apollo_beta, apollo_inputs))

  P <- list()

  for (s in 1:length(pi_values)) {
    work_elasticities <- list(
      Theta = Theta[[s]],
      Phi = Phi[[s]],
      thw = thw[[s]]
    )

    times_elasticities <- list(
      th1 = th1[[s]],
      th2 = th2[[s]]
    )

    sig <- list(
      sig1 = sig1[[s]],
      sig2 = sig2[[s]],
      sig3 = sig3[[s]]
    )

    rho <- list(
      rho12 = rho12[[s]],
      rho13 = rho13[[s]],
      rho23 = rho23[[s]]
    )

    mtuem_settings <- list(
      work_times = c("Tw"),
      free_times = c("Tf1", "Tf2"),
      work_elasticities = work_elasticities,
      times_elasticities = times_elasticities,
      sig = sig,
      rho = rho,
      Tc = Tc, Ec = EcI, w = w, tau = 168,
      componentName = paste0("class_", s)
    )

    P[[paste0("class_", s)]] <- mtuem_likelihood(mtuem_settings, functionality)
  }

  lc_settings <- list(inClassProb = P, classProb = pi_values)
  P[["model"]] <- apollo_lc(lc_settings, apollo_inputs, functionality)
  P <- apollo_prepareProb(P, apollo_inputs, functionality)
  return(P)
}

### Estimate using customMultiStart
multi_starts <- customMultiStart(
  apollo_beta,
  apollo_fixed,
  apollo_probabilities,
  apollo_inputs,
  customMultistart_settings = list(
    apolloBetaMax = apollo_beta + 1,
    apolloBetaMin = apollo_beta - 1,
    nCandidates = 20
  ),
  estimate_settings = est_set,
  first_em = TRUE,
  em_iter_max = 3,
  verbose = FALSE,
  non_saddle = TRUE,
  normalization = list(
    normalization = "theta_phi",
    work_elasticities = c("Theta", "Phi", "thw"),
    times_elasticities = c("th1", "th2"),
    goods_elasticities = c(),
    sig = c("sig1", "sig2", "sig3"),
    rho = c("rho12", "rho13", "rho23")
  ),
  nClass = 2
)

apollo_modelOutput(multi_starts$best_model)

### Predictions
pred <- apollo_prediction(multi_starts$best_model, apollo_probabilities, apollo_inputs)
head(pred[["model"]])
```
