---
title: "Getting Started with mtuem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mtuem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(mtuem)
```

## Introduction

The **mtuem** package provides likelihood functions for estimating Microeconomic Time-Use-Expenditure Models (MTUEM) within the [Apollo](https://www.apollochoicemodelling.com/) choice modelling framework. The approach follows Jara-Diaz et al. (2008), which derives optimal time and expenditure allocations from the maximization of a Cobb-Douglas utility function subject to time and monetary budget constraints.

### Key Concepts

- **Committed time (Tc)**: Time that cannot be freely allocated (sleep, meals, etc.).
- **Committed expenses (Ec)**: Expenses that must be paid regardless of choices.
- **Work time (Tw)**: Time allocated to paid work, generating income at wage rate w.
- **Free time**: Time that can be allocated to leisure activities.
- **Free goods**: Goods that can be purchased with discretionary income.

### Two Parameterizations

The package offers two mathematically equivalent parameterizations:

1. **Theta-Phi** (`mtuem_likelihood`): Uses parameters `Theta`, `Phi`, and `thw`. Requires fixing `Theta = 1` for identification.
2. **Alpha-Beta** (`mtuem_ab_likelihood`): Uses parameters `alpha` and `beta`. No fixed parameters needed.

Both produce identical predictions. Choose based on which parameterization your literature or research group prefers.

## Data Requirements

Your dataset must contain at minimum:

| Variable | Description |
|---|---|
| `Tw` | Work time (hours per week) |
| `Tc` | Committed time (hours per week) |
| `Ec` | Committed expenses (currency per week) |
| `w` | Wage rate (currency per hour) |
| Individual ID | Unique identifier for each observation |

For multi-equation models, add columns for each free time (e.g., `Tf1`, `Tf2`) and free good (e.g., `Ef1`, `Ef2`).

## Step-by-Step Model Formulation

Every MTUEM model in Apollo follows this structure:

### Step 1: Initialise Apollo

```r
library(mtuem)
library(apollo)

apollo_initialise()
apollo_control <- list(
  modelName = "my-model",
  modelDescr = "Description of my model",
  indivID = "PeID"
)
```

### Step 2: Load and Filter Data

```r
database <- arrow::read_parquet("data/maed.parquet.gzip")
database <- database[database$EcI > 0, ]
```

### Step 3: Define Parameters

```r
apollo_beta <- c(Theta = 1, Phi = 1, thw = 0)
apollo_fixed <- c("Theta")  # Theta fixed for identification
```

### Step 4: Validate Inputs

```r
apollo_inputs <- apollo_validateInputs()
```

### Step 5: Define the Probability Function

This is where you configure the model structure:

```r
apollo_probabilities <- function(apollo_beta, apollo_inputs, functionality = "estimate") {
  apollo_attach(apollo_beta, apollo_inputs)
  on.exit(apollo_detach(apollo_beta, apollo_inputs))

  P <- list()

  # Define work elasticities (required)
  work_elasticities <- list(Theta = Theta, Phi = Phi, thw = thw)

  # Define free time elasticities (optional)
  times_elasticities <- list(theta_1 = theta_1)

  # Define free goods elasticities (optional)
  goods_elasticities <- list(phi_1 = phi_1)

  # Define goods costs (optional, default = 1)
  goods_cost <- list(Ef1 = 1)

  # Configure the MTUEM
  mtuem_settings <- list(
    work_times = c("Tw"),
    free_times = c("Tf1"),
    free_goods = c("Ef1"),
    goods_cost = goods_cost,
    work_elasticities = work_elasticities,
    times_elasticities = times_elasticities,
    goods_elasticities = goods_elasticities,
    Tc = Tc, Ec = EcI, w = w, tau = 168
  )

  P[["model"]] <- mtuem_likelihood(mtuem_settings, functionality)
  P <- apollo_prepareProb(P, apollo_inputs, functionality)
  return(P)
}
```

### Step 6: Estimate

```r
model <- apollo_estimate(apollo_beta, apollo_fixed,
                          apollo_probabilities, apollo_inputs)
apollo_modelOutput(model)
```

## The "Leave One Out" Rule

For model identification, you must leave out at least one free time and one free good category. The omitted categories are recovered as residuals:

- **Omitted time** = tau - Tc - Tw - sum(free_times)
- **Omitted good** = w * Tw - Ec - sum(free_goods)

For example, if you have three free time activities (leisure, social, hobbies), include only two in `free_times`. The third is the residual.

## Error Structure Specifications

The likelihood assumes normally distributed errors around the optimal allocations. Choose one of three approaches:

### 1. Inferred Covariance (Single Equation)

Do not specify `sig` or `rho`. The error variance is inferred from residuals.

```r
mtuem_settings <- list(
  work_times = c("Tw"),
  work_elasticities = list(Theta = Theta, Phi = Phi, thw = thw),
  Tc = Tc, Ec = EcI, w = w, tau = 168
  # No sig, no rho
)
```

### 2. Explicit Sigma, Inferred Correlations (Multi-Equation)

Specify `sig` parameters (one per equation). Correlations are inferred from residuals.

```r
mtuem_settings <- list(
  work_times = c("Tw"),
  free_times = c("Tf1"),
  free_goods = c("Ef1"),
  work_elasticities = list(Theta = Theta, Phi = Phi, thw = thw),
  times_elasticities = list(theta_1 = theta_1),
  goods_elasticities = list(phi_1 = phi_1),
  sig = list(sig_1 = sig_1, sig_2 = sig_2, sig_3 = sig_3),
  Tc = Tc, Ec = EcI, w = w, tau = 168
  # No rho
)
```

### 3. Full Covariance Estimation

Specify both `sig` and `rho` parameters. For n equations, you need n sigma parameters and n*(n-1)/2 rho parameters.

For 3 equations, the rho parameters are ordered horizontally across the upper triangle:

```
rho = [rho_12, rho_13, rho_23]
```

This corresponds to the correlation matrix:

```
| 1      rho_12  rho_13 |
| rho_12  1      rho_23 |
| rho_13  rho_23  1     |
```

```r
mtuem_settings <- list(
  work_times = c("Tw"),
  free_times = c("Tf1"),
  free_goods = c("Ef1"),
  work_elasticities = list(Theta = Theta, Phi = Phi, thw = thw),
  times_elasticities = list(theta_1 = theta_1),
  goods_elasticities = list(phi_1 = phi_1),
  sig = list(sig_1 = sig_1, sig_2 = sig_2, sig_3 = sig_3),
  rho = list(rho_12 = rho_12, rho_13 = rho_13, rho_23 = rho_23),
  Tc = Tc, Ec = EcI, w = w, tau = 168
)
```

## Retrieving the Covariance Matrix

After estimation, use the `mtuem_get_corr` function:

```r
result <- mtuem_get_corr(model, apollo_probabilities, apollo_inputs,
                          vars = c("Tw", "Tf1", "Ef1"))
result$covar  # Covariance matrix
result$corr   # Correlation matrix
```

Alternatively, call the likelihood directly with the `get_covar` functionality:

```r
covar_result <- apollo_probabilities(apollo_beta, apollo_inputs,
                                      functionality = "get_covar")
```

## Latent Class Models

For latent class models, use the `customMultiStart` function which handles EM initialization and multi-start optimization.

### Structure

1. Define `apollo_lcPars` to map parameters to classes and compute class probabilities.
2. Loop over classes in `apollo_probabilities`, computing the likelihood for each.
3. Combine using `apollo_lc`.
4. Estimate with `customMultiStart`.

### Example: 1 Class, 1 Equation

```r
apollo_lcPars <- function(apollo_beta, apollo_inputs) {
  lcpars <- list()
  lcpars[["Phi"]] <- list(Phi_1)
  lcpars[["thw"]] <- list(thw_1)

  V <- list()
  V[["class_1"]] <- asc_1

  classAlloc_settings <- list(
    classes = c(class_1 = 1),
    utilities = V
  )
  lcpars[["pi_values"]] <- apollo_classAlloc(classAlloc_settings)
  return(lcpars)
}

apollo_probabilities <- function(apollo_beta, apollo_inputs, functionality = "estimate") {
  apollo_attach(apollo_beta, apollo_inputs)
  on.exit(apollo_detach(apollo_beta, apollo_inputs))

  P <- list()

  for (s in 1:length(pi_values)) {
    work_elasticities <- list(
      Theta = 1,
      Phi = Phi[[s]],
      thw = thw[[s]]
    )

    mtuem_settings <- list(
      work_times = c("Tw"),
      work_elasticities = work_elasticities,
      Tc = Tc, Ec = EcI, w = w, tau = 168,
      componentName = paste0("class_", s)
    )
    P[[paste0("class_", s)]] <- mtuem_likelihood(mtuem_settings, functionality)
  }

  lc_settings <- list(inClassProb = P, classProb = pi_values)
  P[["model"]] <- apollo_lc(lc_settings, apollo_inputs, functionality)
  P <- apollo_prepareProb(P, apollo_inputs, functionality)
  return(P)
}

multi_starts <- customMultiStart(
  apollo_beta, apollo_fixed,
  apollo_probabilities, apollo_inputs,
  first_em = TRUE,
  em_iter_max = 3,
  nClass = 1,
  normalization = list(
    normalization = "theta_phi",
    work_elasticities = c("Theta", "Phi", "thw"),
    times_elasticities = c(),
    goods_elasticities = c(),
    sig = c(),
    rho = c()
  )
)
```

## Post-Estimation: Values of Time

Compute the Value of Leisure Time (VoL) and Value of Time Assigned to Work (VTAW):

```r
pred <- apollo_prediction(model, apollo_probabilities, apollo_inputs)

Tw <- pred$Tw
w  <- database$w
Tc <- database$Tc
Ec <- database$Ec

cteVoL  <- mean((w * Tw - Ec) / (168 - Tw - Tc))
cteVTAW <- mean((w * Tw - Ec) / Tw)

delta <- apollo_deltaMethod(model, list(
  expression = c(
    VoL  = paste0(cteVoL,  "*Theta/Phi"),
    VTAW = paste0(cteVTAW, "*thw/Phi")
  )
))
delta
```

## Troubleshooting

### Model does not converge

- Check that `Theta` is fixed to 1 (Theta-Phi parameterization).
- Ensure the "leave one out" rule is satisfied for free times and goods.
- Try `customMultiStart` with more candidates.
- For latent class models, always use `first_em = TRUE`.

### Negative standard deviations

- Add lower bounds to sigma parameters using Apollo's parameter bounds.
- Use the `normalization` argument in `customMultiStart` to enforce positivity.

### "Leave one out" confusion

If you have K free time activities, include only K-1 in `free_times`. The model recovers the K-th as the residual from the time budget. The same applies to free goods.

### Correlation matrix not positive definite

- When specifying `rho` parameters, ensure they produce a valid correlation matrix.
- Start rho parameters at 0.
- Consider using the inferred correlation approach (specify `sig` but not `rho`).

## References

Jara-Diaz, S. R., Munizaga, M. A., Green, P. E., & Warburton, R. F. (2008). Estimating the value of travel time and of travel time flexibility in the case of workers. *Transportation Research Part B: Methodological*, 42(4), 395-409.

Bunch, D. S., Hamilton, J. H., & Gu, Y. (1993). A Gauss-Newton algorithm for maximum likelihood estimation of nonlinear models with nonsmooth moment conditions. *ACM Transactions on Mathematical Software*, 19(2), 151-179.

Hess, S., & Palma, D. (2019). Apollo: a flexible, powerful and customisable freeware package for choice model estimation and application. *Journal of Choice Modelling*, 32, 100170.
